Incremental Forward Reasoning for White-Box~Proof~Search

### 2025-02-16 paper.tex
Remove a confusing formulation
Minor edits

In many proof assistants, simple automation tactics based on tableau-style tree search play a substantial role in practice: Isabelle's (TODO ref) and Rocq's (TODO ref) auto, ACL2's (TODO ref) waterfall, PVS's (TODO ref) grind, Lean's (TODO ref) Aesop (TODO ref), etc.
These tactics operate on \emph{goals} $Γ ⊢ A$, where the \emph{local context} $Γ$ is a list of assumptions and $A$ is the proposition to be proved.
Faced with an initial goal, the tactics try various user-specified \emph{rules}, which correspond to admissible inferences in the prover's logic.
A successful rule reduces the current goal to zero or more subgoals, which are solved recursively until either a full proof is found or no rule applies to any open goal or the search times out.

@@

However, Aesop's (and ACL2's) support for forward rules was, until recently, quite basic.
In particular, each forward rule was tried on each goal separately (subject to pre-filtering by a discrimination tree (TODO ref) index).
This is wasteful since in practice, many rules produce subgoals with a similar---or even the same---local context as the initial goal.
For example, the ∧-introduction rule affects only the goal's target, not its context.
Still, all forward rules applied to the initial $Γ ⊢ A ∧ B$ would be tried again on the subgoals $Γ ⊢ A$ and $Γ ⊢ B$, with predictably little effect.

In this paper, we present an alternative way to apply forward rules that addresses this issue.
The central idea is to remember, for each goal $Γ ⊢ A$, which forward rules can be partially applied to which sets of hypotheses.
When a non-forward rule then generates subgoals $Δ₁ ⊢ A₁$, \dots, $Δₙ ⊢ Aₙ$, the new local contexts $Δᵢ$ are likely similar to $Γ$.
Hence it is generally cheaper (and often much cheaper) to start with the partial matches for $Γ$ and update them for each of the $Δᵢ$.
The update first deletes partial matches that include hypotheses no longer present in $Δᵢ$.
Afterwards it adds new partial matches for the hypotheses that are new in $Δᵢ$, extending the existing partial matches for a rule if the new hypothesis is compatible with them.
Whenever this process produces a full match---i.e., all premises of a forward rule are satisfied by a set of hypothesis in $Δᵢ$---the rule is applied and adds a corresponding new hypothesis to $Δᵢ$.
Our main contribution is a data structure, the \emph{partial match state}, that organises the partial matches in such a way that the update can be done efficiently.

In addition to supporting such \emph{incremental} forward reasoning, the partial match state satisfies two other desiderata:
\begin{itemize}
  \item It is \emph{persistent}, i.e.\ partial match states for parent and child goals share much of their structure.
        This is desirable since Aesop supports search strategies other than depth-first search and must therefore keep all goals encountered during the search, along with their partial match states, in memory.
  \item It reduces the number of \emph{unifications} performed when matching hypotheses against rule premises.
        In dependent type theory, unification happens up to definitional equality, which makes it quite expensive.
        We therefore want to avoid redundant unifications.
\end{itemize}

Our evaluation (Sec.~\ref{sec:evaluation}) shows that this new implementation of forward reasoning in Aesop substantially outperforms the previous, naive implementation in some (though not all) cases.
The implementation, the benchmarks we used and the full evaluation data are available in a supplement to this paper (TODO ref).
Mainline Aesop already uses the new implementation by default (TODO true?).

@@

If $h : A$ is a hypothesis in the local context of a goal and $A$ unifies with the type $B$ of slot $i$, we say that $h$ \emph{matches} $i$.
Unification induces a substitution $σ$ such that $A[σ]$ is definitionally equal to $B[σ]$; we call this substitution $\sub_{i}(h)$.
For now we do not specify exactly what \enquote{unifies} means; see Sec.~\ref{sec:implementation}.

@@

Instead, we bring all instantiations into a normal form which we call \emph{reducible proof-irrelevant normal form} (RPINF).
This is the usual normal form of expressions with respect to Lean's notion of computation, with two modifications.
First, only \emph{reducible} defined constants are unfolded, i.e.\ replaced with their definiendum.
This reflects our choice to perform unification at reducible transparency.
Second, parts of the original expression that are proofs (in the sense that their types are members of the universe $\Prop$ of propositions) are not normalised.
This reflects the fact that Lean's type theory definitionally equates any two proofs of the same proposition, so there is no need to check whether they are equal.
When computing the RPINF, we also mark all proofs occurring in it with some special \enquote{metadata}.

@@

\paragraph{ACL2.}
Among white-box proof automation tactics, only ACL2's \enquote{waterfall} (TODO ref) has special support for forward reasoning.
(Some such tactics, e.g.\ Coq's auto, can execute arbitrary tactics including forward rules, but make no special provisions for them.)
It uses essentially Aesop's naive algorithm, i.e.\ the local context of each goal is saturated independently without reusing partial results from other goals.
ACL2 additionally supports \emph{trigger terms}, which allow forward rules to be selected based on the presence of a particular term anywhere in the goal, rather than based on particular hypotheses.
Aesop implements a similar mechanism, \emph{rule patterns}, not described here due to space constraints.

@@

We have presented a technique that integrates forward reasoning into a tree-based proof search procedure supporting arbitrary proof rules.
We avoid redundant reasoning steps by caching partial forward rule applications in a data structure, the \emph{forward match state}, that enables efficient updates.
The technique is independent of the search strategy and logic used.
Our evaluation shows a substantial speedup on some synthetic benchmarks, relative to Aesop's previous implementation.

\end{document}

>>>>

In many proof assistants, simple automation tactics based on tableau-style tree search play a substantial role in practice: Isabelle's (TODO ref) and Rocq's (TODO ref) auto, ACL2's (TODO ref) waterfall, PVS's (TODO ref) grind, Lean's (TODO ref) Aesop (TODO ref), etc.
These tactics operate on \emph{goals} $Γ ⊢ A$, where the \emph{context} $Γ$ is a list of assumptions and $A$ is the proposition to be proved.
Faced with an initial goal, the tactics try various user-specified \emph{rules}, which correspond to admissible inferences in the prover's logic.
A successful rule reduces the current goal to zero or more subgoals, which are solved recursively until either a full proof is found or no rule applies to any open goal or the search times out.

@@

However, Aesop's (and ACL2's) support for forward rules was, until recently, quite basic.
In particular, each forward rule was tried on each goal separately (subject to pre-filtering by a discrimination tree (TODO ref) index).
This is wasteful since in practice, many rules produce subgoals with a similar---or even the same---context as the initial goal.
For example, the ∧-introduction rule affects only the goal's target, not its context.
Still, all forward rules applied to the initial $Γ ⊢ A ∧ B$ would be tried again on the subgoals $Γ ⊢ A$ and $Γ ⊢ B$, with predictably little effect.

In this paper, we present an alternative way to apply forward rules that addresses this issue.
The central idea is to remember, for each goal $Γ ⊢ A$, which forward rules can be partially applied to which sets of hypotheses.
When a non-forward rule then generates subgoals $Δ₁ ⊢ A₁$, \dots, $Δₙ ⊢ Aₙ$, the new contexts $Δᵢ$ are likely similar to $Γ$.
Hence it is generally cheaper (and often much cheaper) to start with the partial matches for $Γ$ and update them for each of the $Δᵢ$.
The update first deletes partial matches that include hypotheses no longer present in $Δᵢ$.
Afterwards it adds new partial matches for the hypotheses that are new in $Δᵢ$, extending the existing partial matches for a rule if the new hypothesis is compatible with them.
Whenever this process produces a full match---i.e., all premises of a forward rule are satisfied by a set of hypothesis in $Δᵢ$---the rule is applied and adds a corresponding new hypothesis to $Δᵢ$.
Our main contribution is a data structure, the \emph{partial match state}, that organises the partial matches in such a way that the update can be done efficiently.

In addition to this \emph{incrementality}, the partial match state satisfies two other desiderata:
\begin{itemize}
  \item It is \emph{persistent}, i.e.\ partial match states for parent and child goals share much of their structure.
        This is desirable since Aesop supports search strategies other than depth-first search and must therefore keep all goals encountered during the search, along with their partial match states, in memory.
  \item It reduces the number of \emph{unifications} performed when matching hypotheses against rule premises.
        In dependent type theory, unification happens up to definitional equality, which makes it quite expensive.
        We therefore want to avoid redundant unifications.
\end{itemize}

Our evaluation (Sec.~\ref{sec:evaluation}) shows that this new implementation of forward reasoning in Aesop substantially outperforms the previous, naive implementation on some (though not all) synthetic benchmarks.
The implementation, benchmarks and evaluation data are available in a supplement to this paper (TODO ref).
Mainline Aesop already uses the new implementation by default (TODO true?).

@@

If $h : A$ is a hypothesis in the context of a goal and $A$ unifies with the type $B$ of slot $i$, we say that $h$ \emph{matches} $i$.
Unification induces a substitution $σ$ such that $A[σ]$ is definitionally equal to $B[σ]$; we call this substitution $\sub_{i}(h)$.
For now we do not specify exactly what \enquote{unifies} means; see Sec.~\ref{sec:implementation}.

@@

Instead, we bring all instantiations into a normal form which we call \emph{reducible proof-irrelevant normal form} (RPINF).
This is the usual normal form of expressions with respect to Lean's notion of computation, with two modifications.
First, only \emph{reducible} defined constants are unfolded.
This reflects our choice to perform unification at reducible transparency.
Second, parts of the original expression that are proofs (in the sense that their types are members of the universe $\Prop$ of propositions) are not normalised.
This reflects the fact that Lean's type theory definitionally equates any two proofs of the same proposition, so there is no need to check whether they are equal.
When computing the RPINF, we also mark all proofs occurring in it with some special \enquote{metadata}.

@@

\paragraph{ACL2.}
Among white-box proof automation tactics, only ACL2's \enquote{waterfall} (TODO ref) has special support for forward reasoning.
(Some such tactics, e.g.\ Coq's auto, can execute arbitrary tactics including forward rules, but make no special provisions for them.)
It uses essentially Aesop's naive algorithm, i.e.\ the context of each goal is saturated independently without reusing partial results from other goals.
ACL2 additionally supports \emph{trigger terms}, which allow forward rules to be selected based on the presence of a particular term anywhere in the goal, rather than based on particular hypotheses.
Aesop implements a similar mechanism, \emph{rule patterns}, not described here due to space constraints.

@@

We have presented a technique that integrates forward reasoning into a tree-based proof search procedure supporting arbitrary proof rules.
We avoid redundant reasoning steps by caching partial forward rule applications in a data structure, the \emph{forward match state}, that enables efficient updates.
The technique is independent of the search strategy and logic used.
Our evaluation shows a substantial speedup on some synthetic benchmarks, relative to Aesop's previous implementation.
\end{document}
